---
title: "Mass Shootings In America"
output: html_document
---

```{r Part One:Summary, echo=FALSE, message=FALSE, warning=FALSE}

#Loading Libraries
library(devtools)
library (dplyr)
library (knitr)
library (plotly)

#Creates a dataframe containing information about the shootings and adding another column containing the total number of shootings (injured+killed)
shootings <- read.csv("./Mass-Shooting-Data.csv",stringsAsFactors = FALSE) %>% mutate ( total_shootings = injured + killed)

#Creates a summary about the shootings which occurred 
shooting.summary <- summarise(shootings, injured = sum(injured), killed = sum (killed) , total_shootings = sum (total_shootings))

#Making a dataframe containing the shootings for each city in descending order (highest to lowest)
shootings.for.each.city <- group_by (shootings,city) %>%  summarise(total_shootings=sum(total_shootings)) %>% arrange(-total_shootings)

#Finding the cities with the maximum shootings

#Setting variable for top 'n' cities with the highest number of shootings 
#This is done so that the value can easily be changed
n.cities <- 5

#Finding the cities with the highest number of shootings
#Since the dataframe is already arranged such that rows are sorted in descending order of total shootings we select rows from 1 to 'n'
cities.with.highest.shootings <- shootings.for.each.city [1:n.cities, ]

#The intended statement for the markdown should look like 
#"The cities with the most amount of shootings were (citynames) with (cityvalues) shootings respectively"

#Create a string of the city names together
city.names.highest.shootings <- paste(cities.with.highest.shootings$city,collapse = ", ")

#Create a string of the values together
city.values.highest.shootings <- paste (cities.with.highest.shootings$total_shootings,collapse = ", ")

#Find the starting and end dates for when was the data collected

#Arrange the dates from last date to first date for shootings 
shootings <- shootings[order(as.Date(shootings$date, format="%d/%b/%y")),]

#Create a vector of the dates column from the Shootings dataframe
date.collection <- shootings$date

#Select the first date (since the vector is arranged such that the last date is first..the vector index to select would be the last)
first.date <- date.collection[length(date.collection)]

#Select the last date (since the vector is arranged such that the last date is first..the vector index to select would be the first)
last.date <-date.collection[1]
```

#Introduction

---

> This article presents information about Mass Shootings in America. The data about these mass shootings have been collected through crowdsourcing from ***`r first.date` to `r last.date`***. There were a total of ***`r shooting.summary$total_shootings` shootings*** which caused  ***`r shooting.summary$injured` people to get injured *** and ***`r shooting.summary$killed` people's deaths.*** The cities with the most amount of shootings were ***`r city.names.highest.shootings` with `r city.values.highest.shootings` shootings.***

---

##Summary Tables

```{r Part One:Summary Tables, results = 'asis', echo=FALSE}

#Transposing the dataframe for better presenation of the data in the HTML document
shooting.summary <- t (shooting.summary)

#Renaming the column and row names for better table presentation in the HTML document 
colnames(shooting.summary)[1] <- "Total"
row.names(shooting.summary)[1] <- "Injuries"
row.names(shooting.summary)[2] <- "Killed"
row.names(shooting.summary)[3] <- "Shootings"

#Creating a table containing a summary of shootings in USA
kable(shooting.summary, digits=2, caption = "Summary of Shootings in USA")

#Renaming the column names so they look better in the table
colnames(cities.with.highest.shootings)[1] <- "City Name"
colnames(cities.with.highest.shootings)[2] <- "Total Shootings"

#Creating a table containing Cities with Highest Amount of Killings in USA
kable(cities.with.highest.shootings, digits = 2,  caption = "Cities with Highest Amount of Killings in USA")

```

```{r Part Two: Incident Description ,echo=FALSE, message=FALSE, warning=FALSE}
#I wish to find out the incident with the maximum killings on a particular date
#In my case study summary I provide information about that incident along with other useful stats

#Setting the date
shooting.date <- '15-Oct-16'

#Getting all the data about the shootings on that date
shooting.info <- filter (shootings, date == shooting.date) 

#Finding out the most fatal incident of the day by filtering out first by maximum amount of people impacted (total_shootings) then by amount of people killed and then by people injured and then selecting one incident from the data since they all will have the same statistics after this filtering procedure
#Note: It is ensureed that only one incident is selected as there could be a possibility that when the data is updated there could be two incidents with matching statistics
shooting.info.max <- (filter (shooting.info, total_shootings == max (total_shootings)) %>% filter(killed == max (killed)) %>% filter(injured == max (injured) ))[1,]

#Dataframe containing information about other shootings on that date
other.shootings.on.date <-filter (shooting.info, total_shootings != max (total_shootings))
#Number of other shootings on that date
other.shootings.count <- nrow(other.shootings.on.date)
#Cities in which the other shootings occurred
other.shootings.cities <- paste (other.shootings.on.date$city, collapse=", ")


#Finding out the most fatal and least fatal incidents so as to provide more interesting information about the shooting being studied to the user

#THe most fatal incident is found by filtering out first by maximum amount of people impacted (total_shootings) then by amount of people killed and then by people injured and then selecting one incident from the data since they all will have the same statistics after this filtering procedure
#Note: It is ensureed that only one incident is selected as there could be a possibility that when the data is updated there could be two incidents with matching statistics
shooting.max <- (filter (shootings, total_shootings == max (total_shootings)) %>% filter(killed == max (killed)) %>% filter(injured == max (injured) ))[1,]

#The least fatal incident is found by filtering out first by minimum amount of people impacted (total_shootings) then by amount of people killed and then by people injured and then selecting one incident from the data since they all will have the same statistics after this filtering procedure
#Note: It is ensureed that only one incident is selected as there could be a possibility that when the data is updated there could be two incidents with matching statistics
shooting.min <- (filter (shootings, total_shootings == min (total_shootings)) %>% filter(killed == min (killed)) %>%  filter(injured == min (injured)))[1,]

#Finding the difference between the most fatal and the incident being studied
diff.with.most.fatal <- shooting.max$total_shootings - shooting.info.max$total_shootings

#Finding the difference between the least fatal and the incident being studied
diff.with.least.fatal <- shooting.info.max$total_shootings - shooting.min$total_shootings 

```
---

#Single Incident Study

---
> ##About
**On `r shooting.date` the most fatal shooting was in `r shooting.info.max$city`, `r shooting.info.max$state`.
It impacted `r shooting.info.max$total_shootings` people - `r shooting.info.max$injured` injured and `r shooting.info.max$killed` killed.** 


> ##Further Information
* There were `r other.shootings.count` other shootings on that day - `r other.shootings.cities`.
* `r diff.with.most.fatal` less than the most fatal incident of the year (`r shooting.max$total_shootings` impacted) 
*  `r diff.with.least.fatal` more than the least fatal incident of the year (`r shooting.min$total_shootings` impacted).


```{r Part Three: Interactive Map ,echo=FALSE, message=FALSE, warning=FALSE}
#Creating Dataframe for Data Visualization via the form of an Interactive Map
shootings.visualization <- group_by (shootings,city,state,lng,lat) %>%  summarise(total_shootings=sum(total_shootings), killed =sum (killed), injured =sum(injured)) %>% arrange(-total_shootings) %>% ungroup ()

#Create a column which contains the information to show on hover over a location
shootings.visualization$hover <- with(shootings.visualization, paste(city,",",state ,"<br>" , "Injured:", injured,"<br>", "Killed:", killed, "<br>","People Shot",total_shootings))

# Geo styling
g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showland = TRUE,
  landcolor = toRGB("black"),
  subunitcolor = toRGB("grey85"),
  countrycolor = toRGB("grey85"),
  countrywidth = 0.5,
  subunitwidth = 0.5
)

#Color of the Bubble
bubble.color <- "red"

map <- plot_geo(shootings.visualization, locationmode = 'USA-states', sizes = c(1, 250)) %>% add_markers(x = ~lng, y = ~lat, size = ~total_shootings, color = bubble.color, text=~hover)%>% layout(title = '2016 Shootings', geo = g)
```
#Interactive Map

`r map`

>*This is a bubble map displaying information about the intensity of shootings in each city of USA - the size of the bubble is decided by the total number of shootings for a city).* 

>*From the map above we can notice that eastern states of USA experience more shootings that the western states.*
```{r Part Four: Interactive Plot ,echo=FALSE, message=FALSE, warning=FALSE}

#DataFrame created which contains the state, total shootings and total kills for that state and then a column has been added which contains the death percentage for the shootings for each state
data.interactive.plot <- group_by (shootings,state) %>%  summarise(total_shootings=sum(total_shootings), killed = sum (killed), injured=sum(injured)) %>% ungroup () %>% mutate (death_percentage = ((killed/total_shootings)*100))


#shootings.visualization$hover <- with(shootings.visualization, paste(city,",",state ,"<br>" , "Injured:", injured,"<br>", "Killed:", killed, "<br>","Shootings",total_shootings))

data.interactive.plot$hover <- with(data.interactive.plot, paste(state ,"<br>" , "Injured:", injured,"<br>", "Killed:", killed, "<br>","People Shot:",total_shootings,"<br>","Death Percentage:",death_percentage))

font.style <- list(
  family = "Courier New, monospace",
  size = 18,
  color = "#7f7f7f"
)
x <- list(
  title = "Death Percentage for each state",
  titlefont = font.style
)
y <- list(
  title = "Total Shootings for each state",
  titlefont = font.style
)

interactive.plot <- plot_ly(   data.interactive.plot, x=~death_percentage, y=~total_shootings, text=~hover, type ='scatter', mode ='markers', color= ~death_percentage) %>% layout(xaxis=x,yaxis=y,title="Death Rate (in %) for Total Shootings in each State")

```

#Interactive Plot

>I was really interested in finding out the death rate for the amount of people shot per state as I wanted to investigate if a higher amount of people shot value meant a higher death rate. 

*I compiled together the following data for each state - total shootings,total killed and then calculated the death percentage for the shootings and created an interactive scatter plot to display this information.*

***Hover a value to find out more information such as the state name***

`r interactive.plot`


>**As can be seen from the plot above a higher amount of people shot does not always mean a higher death percentage. Infact in the state of Washington there were a small number of people shot compared to other shootings around USA. However the death percentage was still high. Whereas in the state of Florida which has the maximum amount of people shot the death percentage is comparitvely low compared to Washington **
